Annotated example of a spatially varying effect with sdmTMB

### 2021-11-27 doc/appendix-snow-sdmTMB.Rmd
change theme, rename projection...

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
library(sf)
```

@@

Here, X and Y are coordinates in Albers projection for North America and divided by 100000 to give units of 100 km. 
This projection was achieved with the following crs code using `sf::st_transform(crs = proj)`.

```{r proj, echo=TRUE}
proj <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 
+datum=NAD83 +units=m +no_defs"
```

@@

```{r explore, fig.asp = 0.3}
ggplot(snow, aes(year, count, colour = nao)) + 
  geom_jitter() + 
  scale_colour_viridis_c() +
  theme_bw() 
```

@@

```{r fit-poisson, echo=TRUE, eval=FALSE}
m0 <- sdmTMB(count ~ 1 + nao + (1|year_f),
             time = "year",
             spatial_varying = ~ 0 + nao,
             spatial = "on", 
             spatiotemporal = "IID",
             family = poisson(link = "log"),
             data = snow, mesh = mesh)
```

```{r fit-negb1, echo=TRUE, eval=FALSE}
m1 <- sdmTMB(count ~ 1 + nao + (1|year_f),
            time = "year",
            spatial_varying = ~ 0 + nao,            
            spatial = "on", 
            spatiotemporal = "IID",
            family = nbinom1(link = "log"),
            data = snow, mesh = mesh)
```

```{r fit-negb2, echo=TRUE, eval=FALSE}
m2 <- sdmTMB(count ~ 1 + nao + (1|year_f),
            time = "year",
            spatial_varying = ~ 0 + nao,            
            spatial = "on", 
            spatiotemporal = "IID",
            family = nbinom2(link = "log"),
            data = snow, mesh = mesh)
```

@@

```{r fit-negb2-reml, echo=TRUE, eval=FALSE}
m <- sdmTMB(count ~ 1 + nao + (1|year_f),
            time = "year",
            spatial_varying = ~ 0 + nao,            
            spatial = "on", 
            spatiotemporal = "IID",
            family = nbinom2(link = "log"),
            reml = TRUE,
            data = snow, mesh = mesh)
```

Print the model to inspect the fit:

```{r print-fit, warning=FALSE, echo=TRUE, message=FALSE}
print(m)
```

@@

We can inspect randomized quantile residuals, although this method of checking residuals is not guaranteed to be accurate.

@@

```{r plot-resids, fig.width = 8}
theme_set(theme_bw() + theme(
  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
  legend.position = "top", legend.justification = c(0, 0), 
  axis.title = element_blank()))

ggplot(pred, aes(X, Y, colour = resid)) +
  geom_point(size = 0.5) +
  facet_wrap(~year) +
  coord_fixed() +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  scale_colour_gradient2(guide = guide_colourbar(title.position = "top"))
```

@@

In order to produce pretty maps of these predictions, we back transform the spatial coordinates into the same units as our original projection and then convert back into an sf object so that the plot axis will appear in latitude and longitude.

```{r proj-p, echo=TRUE, eval=T}
library(sf)

p <- pred %>% mutate(
  X = X * 100000,
  Y = Y * 100000
)

p_proj <- p %>% mutate(x = X, y = Y) %>% st_as_sf(., coords = c("x", "y"), crs = proj)
```

One can also get continental coastlines from the package rnaturalearth and outlines of the great lakes from [http://www.naturalearthdata.com/](http://www.naturalearthdata.com/) and transform them to the correct projection using `sf::st_transform(crs = proj)`.

```{r shapes, warning=FALSE, echo=TRUE, message=FALSE}
library("rnaturalearth")
library("rnaturalearthdata")
coast <- ne_coastline(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(crs = proj)
lakes <- sf::st_read(here::here("owls/ne_10m_lakes"),  quiet = TRUE)
lakes <- lakes[lakes$scalerank==0,] %>% sf::st_transform(crs = proj) # only large lakes
```

@@

```{r plot-map, echo=TRUE, eval=T}
ggplot(data = p_proj) +
  geom_point(data = p_mean, 
     aes(X,Y, colour = nao_effect, size = mean_est_count), alpha = 0.5) +
  geom_sf(data = coast, fill = NA) +
  geom_sf(data = lakes, fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),# adjusts space on sides
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top")) +
  guides(size = "none") + 
  labs(colour = "NAO effect\non Snowy Owl count") +
  theme_bw() + theme(
    legend.key.height = unit(0.3, "cm"),
    legend.position = c(0.25,0.13),
    axis.title = element_blank())
```

@@

```{r plot-all-effects, echo=TRUE, eval=T}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(trans="sqrt", # makes variation in low numbers visible 
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")) +
  labs(colour = "Predicted counts") + 
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-fixed-effects, echo=TRUE, eval=T}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est_non_rf)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(breaks = c(0.03, 0.05, 0.07), 
    guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-spatial-effects, echo=TRUE, eval=T}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X,Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, fill = NA) +
  geom_sf(data = lakes, fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top")) +
  theme_bw() + theme(
    legend.key.height = unit(0.3, "cm"),
    legend.position = c(0.25,0.13),
    axis.title = element_blank())
```

@@

```{r plot-spatiotemporal-effects, echo=TRUE, eval=T}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = epsilon_st), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

>>>>

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
library(sf)
theme_set(ggsidekick::theme_sleek() + theme( # modifications for plotting sf data
  legend.position = "top", legend.justification = c(0, 0), # move legend to top left
  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"), 
  panel.grid.major = element_line(colour = "grey90"), # for lat/lons
  axis.title = element_blank())) # lat/lon is clear from axis text
```

@@

Here, X and Y are coordinates in Albers projection for North America and divided by 100000 to give units of 100 km. 
This projection was achieved with the following crs code using `sf::st_transform(crs = Albers)`.

```{r proj, echo=TRUE}
Albers <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 
+datum=NAD83 +units=m +no_defs"
```

@@

```{r explore, fig.asp = 0.3}
ggplot(snow, aes(year, count, colour = nao)) + 
  geom_jitter() + 
  scale_colour_viridis_c() +
  ggsidekick::theme_sleek()
```

@@

```{r fit-poisson, echo=TRUE, eval=FALSE}
m0 <- sdmTMB(count ~ 1 + nao + (1|year_f),
             spatial_varying = ~ 0 + nao,
             time = "year",
             spatial = "on", 
             spatiotemporal = "IID",
             family = poisson(link = "log"),
             data = snow, mesh = mesh)
```

```{r fit-negb1, echo=TRUE, eval=FALSE}
m1 <- sdmTMB(count ~ 1 + nao + (1|year_f),
            spatial_varying = ~ 0 + nao,  
            time = "year",
            spatial = "on", 
            spatiotemporal = "IID",
            family = nbinom1(link = "log"),
            data = snow, mesh = mesh)
```

```{r fit-negb2, echo=TRUE, eval=FALSE}
m2 <- sdmTMB(count ~ 1 + nao + (1|year_f),
            spatial_varying = ~ 0 + nao,  
            time = "year",
            spatial = "on", 
            spatiotemporal = "IID",
            family = nbinom2(link = "log"),
            data = snow, mesh = mesh)
```

@@

```{r fit-negb2-reml, echo=TRUE, eval=FALSE}
m <- sdmTMB(count ~ 1 + nao + (1|year_f), spatial_varying = ~ 0 + nao,
            time = "year", spatial = "on", spatiotemporal = "IID",
            family = nbinom2(link = "log"),
            reml = TRUE,
            data = snow, mesh = mesh)
```

Print the model to inspect the fit:
```{r check, include = FALSE, cache=FALSE, eval=TRUE}
m 
```

```{r print-fit, warning=FALSE, echo=TRUE, message=FALSE, cache=FALSE}
print(m)
```

@@

\clearpage
We can inspect randomized quantile residuals, although this method is not guaranteed to be accurate.

@@

We can also test for zero-inflation using the `simulate.sdmTMB` function and a couple functions from the DHARMa package.
<!-- If our data was zero-inflated ... maybe make suggestions or even link to an example for how to do a delta model in sdmTMB?-->

```{r zero-test, echo=TRUE, message=FALSE, cache=FALSE}
s_nb2 <- simulate(m, nsim = 500) 
pred_fixed <- m$family$linkinv(predict(m)$est_non_rf)
r_nb2 <- DHARMa::createDHARMa(
  simulatedResponse = s_nb2,
  observedResponse = m$data$count,
  fittedPredictedResponse = pred_fixed
)
DHARMa::testZeroInflation(r_nb2, plot = FALSE)
```

<!-- \clearpage -->
Check the spatiotemporal distribution of the residuals.

```{r plot-resids, fig.width = 8, cache=FALSE}
ggplot(pred, aes(X, Y, colour = resid)) + geom_point(size = 0.5) +
  facet_wrap(~year) + coord_fixed() +
  scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) + 
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

@@

In order to produce pretty maps of these predictions, we back transform the spatial coordinates into the same units as our original projection and then convert back into an sf object so that the plot axis will be in latitude and longitude.

```{r proj-p, echo=TRUE, eval=T}
p <- pred %>% mutate(X = X * 100000, Y = Y * 100000)
p_proj <- p %>% mutate(x = X, y = Y) %>% sf::st_as_sf(coords = c("x", "y"), crs = Albers)
```

We retrieved continental coastlines from the package rnaturalearth and outlines of the great lakes from [http://www.naturalearthdata.com/](http://www.naturalearthdata.com/) and transformed them to the correct projection.

```{r shapes, warning=FALSE, echo=TRUE, message=FALSE}
coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(crs = Albers) 
lakes <- sf::st_read(here::here("owls/ne_10m_lakes"),  quiet = TRUE)
lakes <- lakes[lakes$scalerank==0,] %>% sf::st_transform(crs = Albers) # only large lakes
```

@@

```{r plot-map, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = p_proj) + 
  geom_point(data = p_mean, 
     aes(X,Y, colour = nao_effect, size = mean_est_count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),# adjusts space on sides
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top")) +
  guides(size = "none") + 
  labs(colour = "NAO effect\non Snowy Owl count") +
  theme(legend.key.height = unit(0.3, "cm"),
        legend.position = c(0.1,0.1),
        axis.title = element_blank())
```

@@

```{r plot-all-effects, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = filter(p_proj, year > 1995)) + 
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(trans="sqrt", # makes variation in low numbers visible 
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")) +
  labs(colour = "Predicted counts") + 
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-fixed-effects, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = filter(p_proj, year > 1995)) + 
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est_non_rf)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(breaks = c(0.03, 0.05, 0.07), 
    guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-spatial-effects, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = p_proj) + 
  geom_point(data = p, aes(X,Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top")) +
  theme(legend.key.height = unit(0.3, "cm"),
        legend.position = c(0.1,0.1),
        axis.title = element_blank())
```

@@

```{r plot-spatiotemporal-effects, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = filter(p_proj, year > 1995)) + 
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = epsilon_st), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```
